---
title: "Final Report"
author: "Phuc Nguyen, Joseph Lawson, Emily Gentles"
date: "1/20/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Abstract/Summary - summarize the problem and the main findings of your project.

# Section 1: Introduction - provide an initial description of the data, study in which it was collected and focus of the study/analysis/questions of interest.

  The data used in this analysis is a subsample of the data collected in the National Collaborative Perinatal Project (CPP), a program which enrolled women during pregnancy and followed their pregnancy outcomes. The project collected demographic information such as race and maternal age as well as lifestyle factors such as smoking status. The subsample contains 2380 women and children for whom additional assaying of serum samples was perfromed and the concentration doses for DDE and PCBs was recorded. DDE and PCBs are of particular interest as they are breakdown products of pesticides that bioaccumulate in fatty deposits of organisms. The goal of this analysis is to assess how exposure to DDE and PCBs relates to the risk of premature delivery. 

  Many of the variables in this study exhibit a variety of issues such as uncertainty, inflation, and missingness. Uncertainty is introduced into the variable gestational age as calculating the date of conception is not exact because it typically relies on the first day of the mother's last menstrual period. Assuming the mother did not miss any menstrual cycles, the date of conception could be incorrect by 11-21 days meaning that a gestational age of 47, normally considered highly unlikely, could atually represent a gestational age of 44, normally considered as a plausable late term gestation. There also seems to be some general mismanagement of data as gestational ages of 90 weeks appear in the data even though this is almost twice the length of the longest recorded pregnancy. We also see an inflation of zeros in the PCB variables due to the limitations of measurement equipment that, below a certain threshold, rounds values to zero. Missigness is most significantly present for the variable albumin for which less than 10% of women had their levels recorded. Since albumin can be an indicator of liver disease, kidney disease, and possibly gestational diabetes, the missigness might be due to the fact that only women who were perceived to be at risk were tested. In this case, the simple presence of being tested for albumin would likely be more significant than the actual values recorded.

# Section 2: Materials & Methods - describe the statistical approach/approaches used in the analysis, providing references as appropriate; do not present results in this section but just the methods used. You can carry out your analysis with multiple methods.

In order to assess how exposure to DDE and PCBs relates to the risk of premature delivery we decided to investigate how well the concentration doses of DDE and PCBs predict premature delivery. We noticed that there are clinical categories for gestational age with gestational ages of less than 33 weeks corresponding to severely preterm births, gestational ages of less than 38 weeks corresponding to preterm births, and gestational ages of between 38 and 42 corresponding to normal term births. Considering that our interest is in the risk of premature delivery, we decided to disregard gestational ages of greater than 44 weeks which would have correspond to late term births. The structure of these categories lead us to conduct ordinal logistic regression. This method provides useful interpretatiion in terms of risk while using the naturally ordinal structure of the gestational age categories and differentiating between the severity of severly preterm births and preterm births. This structure also addresses the non-normality of the response. 

- Ordinal Logistic Regression with Term, Preterm, and Severely Preterm Gest. Categories
  + Useful interpretation in terms of risk
  + Uses naturally ordinal structure
  + Differentiates between different severities
  + Addresses non-normality of response
- Keep obs with Gest. Age 44 or less
- Impute score data with MICE to check usefulness
- Remove obs with missing PCE value
- Include blood cholesterol/triglyceride levels, as well as center and SES/Lifestyle metrics


# Section 3: Results - describe the results of the analysis approaches described in Section 2. It
may be useful to break this into subsections, such as:
– 3.1 Exploratory Data Analysis
– 3.2 Main Results
– 3.3 Sensitivity Analysis
Make sure to very clearly address the questions of interest!

### EDA

```{r, library, message=FALSE, results=FALSE}
library(tidyverse)
library(mice)
library(MASS)
```

```{r, clean-data}
longnecker_base <- readRDS(file = "Longnecker.rds")

# Keep gestational age less than 45
# Remove 1 row of pcb_* with NA
# Dichotomize gestational age into 4 categories
# Create indicator of albumin
# Change center to categorical
gestational_age_cutoff <- 45
longnecker <- longnecker_base %>%
  filter(gestational_age < gestational_age_cutoff) %>%
  filter(!is.na(pcb_028)) %>%
  mutate(has_albumin = 1 * !is.na(albumin)) %>%
  mutate(center = as.character(center)) %>%
  mutate(gestational_cat = as.factor(ifelse(gestational_age <= 32, 1, 
                                            ifelse(gestational_age <= 37 & gestational_age > 32, 2, 3))))
                                                   #ifelse(gestational_age > 37 & gestational_age <= 41, 3, 4)))))

# Log transform pcb_* after adding a small number to 0 values
pcb_mat <- longnecker %>%
                    dplyr::select(contains("pcb")) %>%
                    data.matrix()
min_pcb <- min(pcb_mat[pcb_mat > 0])
pcb_mat[pcb_mat == 0] <- min_pcb / 2
pcb_mat <- apply(pcb_mat, 2, log)
# Summarize pcb_* with PCA
pcb_pca <- prcomp(pcb_mat, scale = TRUE)
# To see loading of each variable, uncomment the below:
#pcb_pca$rotation[, 1] * (pcb_mat[1, ] - pcb_pca$center)/ pcb_pca$scale
# To see variation explained by each PC, uncomment the below:
# pcb_pca$sdev^2 # == > two PCs are enough
pcb_comps <- pcb_mat %*% pcb_pca$rotation[, c(1,2)]
longnecker["pcb_pca1"] <- pcb_comps[, 1]
longnecker["pcb_pca2"] <- pcb_comps[, 2]

# Log transform these variables
longnecker["log_dde"] <- log(longnecker$dde)
longnecker["log_tri"] <- log(longnecker$triglycerides)
```

```{r}

```


### Ordered logistic regression

```{r, ord}
ord_mod <- polr(gestational_cat ~ race +
                 dde + 
                 pcb_pca1 +
                 has_albumin +
                 poly(maternal_age, 2) + 
                 log_tri +
                 cholesterol +
                 smoking_status +
                 center, 
               data = longnecker,
               Hess=TRUE)
p_ord <- pnorm(abs(coef(summary(ord_mod))[, "t value"]), lower.tail = FALSE) * 2
cbind(summary(ord_mod)$coefficients, p_ord)
```

```{r, diagnostics}
probabilities <- predict(ord_mod, type = "probs")
predictors <- c("log_dde", 
                 "pcb_pca1",
                 "log_tri",
                 "cholesterol",
                 "severe_premature", "premature", "normal"
                )
longnecker %>%
  mutate(severe_premature = log(probabilities[, 1]/(1-probabilities[, 1])),
         premature = log(probabilities[, 2]/(1-probabilities[, 2])),
         normal = log(probabilities[, 3]/(1-probabilities[, 3]))
         ) %>%
  dplyr::select(predictors) %>%
  gather(key = "predictors", value = "predictor.value", 
         -severe_premature, -premature, -normal) %>%
  gather(key = "categories", value = "logit.value", -predictors, -predictor.value) %>%
  ggplot(aes(logit.value, predictor.value))+
  geom_point(size = 0.5, alpha = 0.5) +
  geom_smooth(method = "loess") + 
  theme_bw() + 
  facet_grid(cols = vars(categories), rows = vars(predictors), scales = "free")
```

### Sensitivity Analysis

```{r}
pred <- predict(ord_mod, type = "probs")
predictors <- c("log_dde", 
                 "pcb_pca1",
                 "log_tri",
                 "cholesterol",
                 "severe_premature", "premature", "normal","late"
                )
pred1 <- (pred[,1])
pred1.prob <- sum(pred1)
pred2 <- (pred[,2])
pred2.prob <- sum(pred2)
pred3 <- (pred[,3])
pred3.prob <- sum(pred3)
sum <- pred1.prob + pred2.prob + pred3.prob

pred1.p <- pred1.prob/sum
pred2.p <- pred2.prob/sum
pred3.p <- pred3.prob/sum
pred1.p+pred2.p+pred3.p

l1 <- longnecker$gestational_cat == "1"
l1.s <- sum(l1 == TRUE)
l2 <- longnecker$gestational_cat == "2"
l2.s <- sum(l2 == TRUE)
l3 <- longnecker$gestational_cat == "3"
l3.s <- sum(l3 == TRUE)
suml <- length(longnecker$gestational_cat)

l1.p <- l1.s/suml
l2.p <- l2.s/suml
l3.p <- l3.s/suml

#s <- with(longnecker, summary(ord_mod))
#s
#plot(s, which=1:3, pch=1:3)

install.packages('jtools')
install.packages('ggstance')
library(ggstance)
library(jtools)
# can use to plot coefs from multiple models
plot_summs(ord_mod)
```


```{r, cache=TRUE}
y <- longnecker$gestational_cat
n <- 200
y_hat <- matrix(NA, nrow = length(y), ncol = n)
for (i in 1:n) {
  y_hat[, i] <- apply(probabilities, 1, function(x) {which(rmultinom(1, 1, x) == 1)})
}
count_severe <- apply(y_hat, 2, function(x) {sum(x == 1)})
count_mild <- apply(y_hat, 2, function(x) {sum(x == 2)})
count_normal <- apply(y_hat, 2, function(x) {sum(x == 3)})
sens_data <- data.frame(severe = count_severe, mild = count_mild, normal = count_normal) %>%
  gather(key = "categories", value = "count")
ggplot(sens_data, aes(x = count, fill = categories)) + 
  geom_histogram(bins = 200) +
  geom_vline(xintercept = 69, color = "red", size = 0.5) +
  geom_vline(xintercept = 453, color = "red", size = 0.5) +
  geom_vline(xintercept = 1789, color = "red", size = 0.5)
```


# Section 4: Discussion - comment on advantages and disadvantages of the approach taken and briefly discuss other approaches you could have taken instead.



# Appendix - contains any necessary technical and other material that would be good to include but may make the main part difficult to read, particularly for a non-specialist/scientist. Sections 1-4 should be as accessible as possible to a non-statistician scientist/researcher in the relevant area & should be written in a clear/concise manner.

# Citations
cut offs
https://www.who.int/news-room/fact-sheets/detail/preterm-birth

albumin
https://account.allinahealth.org/library/content/1/3480
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4665611/