---
title: "Final Report"
author: "Phuc Nguyen, Joseph Lawson, Emily Gentles"
date: "1/20/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Abstract/Summary - summarize the problem and the main findings of your project.

# Section 1: Introduction - provide an initial description of the data, study in which it was
collected and focus of the study/analysis/questions of interest.

# Section 2: Materials & Methods - describe the statistical approach/approaches used in the
analysis, providing references as appropriate; do not present results in this section but just the
methods used. You can carry out your analysis with multiple methods.

# Section 3: Results - describe the results of the analysis approaches described in Section 2. It
may be useful to break this into subsections, such as:
– 3.1 Exploratory Data Analysis
– 3.2 Main Results
– 3.3 Sensitivity Analysis
Make sure to very clearly address the questions of interest!

### EDA

```{r, library, message=FALSE, results=FALSE}
library(tidyverse)
library(mice)
library(MASS)
```

```{r, clean-data}
longnecker_base <- readRDS(file = "Longnecker.rds")

# Keep gestational age less than 45
# Remove 1 row of pcb_* with NA
# Dichotomize gestational age into 4 categories
# Create indicator of albumin
# Change center to categorical
gestational_age_cutoff <- 45
longnecker <- longnecker_base %>%
  filter(gestational_age < gestational_age_cutoff) %>%
  filter(!is.na(pcb_028)) %>%
  mutate(has_albumin = 1 * !is.na(albumin)) %>%
  mutate(center = as.character(center)) %>%
  mutate(gestational_cat = as.factor(ifelse(gestational_age <= 32, 1, 
                                            ifelse(gestational_age <= 37 & gestational_age > 32, 2, 3))))
                                                   #ifelse(gestational_age > 37 & gestational_age <= 41, 3, 4)))))

# Log transform pcb_* after adding a small number to 0 values
pcb_mat <- longnecker %>%
                    dplyr::select(contains("pcb")) %>%
                    data.matrix()
min_pcb <- min(pcb_mat[pcb_mat > 0])
pcb_mat[pcb_mat == 0] <- min_pcb / 2
pcb_mat <- apply(pcb_mat, 2, log)
# Summarize pcb_* with PCA
pcb_pca <- prcomp(pcb_mat, scale = TRUE)
# To see loading of each variable, uncomment the below:
#pcb_pca$rotation[, 1] * (pcb_mat[1, ] - pcb_pca$center)/ pcb_pca$scale
# To see variation explained by each PC, uncomment the below:
# pcb_pca$sdev^2 # == > two PCs are enough
pcb_comps <- pcb_mat %*% pcb_pca$rotation[, c(1,2)]
longnecker["pcb_pca1"] <- pcb_comps[, 1]
longnecker["pcb_pca2"] <- pcb_comps[, 2]

# Log transform these variables
longnecker["log_dde"] <- log(longnecker$dde)
longnecker["log_tri"] <- log(longnecker$triglycerides)
```

```{r}

```


### Ordered logistic regression

```{r, ord}
ord_mod <- polr(gestational_cat ~ race +
                 dde + 
                 pcb_pca1 +
                 has_albumin +
                 poly(maternal_age, 2) + 
                 log_tri +
                 cholesterol +
                 smoking_status +
                 center, 
               data = longnecker,
               Hess=TRUE)
p_ord <- pnorm(abs(coef(summary(ord_mod))[, "t value"]), lower.tail = FALSE) * 2
cbind(summary(ord_mod)$coefficients, p_ord)
```

```{r, diagnostics}
probabilities <- predict(ord_mod, type = "probs")
predictors <- c("log_dde", 
                 "pcb_pca1",
                 "log_tri",
                 "cholesterol",
                 "severe_premature", "premature", "normal"
                )
longnecker %>%
  mutate(severe_premature = log(probabilities[, 1]/(1-probabilities[, 1])),
         premature = log(probabilities[, 2]/(1-probabilities[, 2])),
         normal = log(probabilities[, 3]/(1-probabilities[, 3]))
         ) %>%
  dplyr::select(predictors) %>%
  gather(key = "predictors", value = "predictor.value", 
         -severe_premature, -premature, -normal) %>%
  gather(key = "categories", value = "logit.value", -predictors, -predictor.value) %>%
  ggplot(aes(logit.value, predictor.value))+
  geom_point(size = 0.5, alpha = 0.5) +
  geom_smooth(method = "loess") + 
  theme_bw() + 
  facet_grid(cols = vars(categories), rows = vars(predictors), scales = "free")
```

### Sensitivity Analysis

<<<<<<< HEAD
```{r}
pred <- predict(ord_mod, type = "probs")
predictors <- c("log_dde", 
                 "pcb_pca1",
                 "log_tri",
                 "cholesterol",
                 "severe_premature", "premature", "normal","late"
                )
pred1 <- (pred[,1])
pred1.prob <- sum(pred1)
pred2 <- (pred[,2])
pred2.prob <- sum(pred2)
pred3 <- (pred[,3])
pred3.prob <- sum(pred3)
sum <- pred1.prob + pred2.prob + pred3.prob

pred1.p <- pred1.prob/sum
pred2.p <- pred2.prob/sum
pred3.p <- pred3.prob/sum
pred1.p+pred2.p+pred3.p

l1 <- longnecker$gestational_cat == "1"
l1.s <- sum(l1 == TRUE)
l2 <- longnecker$gestational_cat == "2"
l2.s <- sum(l2 == TRUE)
l3 <- longnecker$gestational_cat == "3"
l3.s <- sum(l3 == TRUE)
suml <- length(longnecker$gestational_cat)

l1.p <- l1.s/suml
l2.p <- l2.s/suml
l3.p <- l3.s/suml

#s <- with(longnecker, summary(ord_mod))
#s
#plot(s, which=1:3, pch=1:3)

install.packages('jtools')
install.packages('ggstance')
library(ggstance)
library(jtools)
# can use to plot coefs from multiple models
plot_summs(ord_mod)
```


=======
```{r, cache=TRUE}
y <- longnecker$gestational_cat
n <- 200
y_hat <- matrix(NA, nrow = length(y), ncol = n)
for (i in 1:n) {
  y_hat[, i] <- apply(probabilities, 1, function(x) {which(rmultinom(1, 1, x) == 1)})
}
count_severe <- apply(y_hat, 2, function(x) {sum(x == 1)})
count_mild <- apply(y_hat, 2, function(x) {sum(x == 2)})
count_normal <- apply(y_hat, 2, function(x) {sum(x == 3)})
sens_data <- data.frame(severe = count_severe, mild = count_mild, normal = count_normal) %>%
  gather(key = "categories", value = "count")
ggplot(sens_data, aes(x = count, fill = categories)) + 
  geom_histogram(bins = 200) +
  geom_vline(xintercept = 69, color = "red", size = 0.5) +
  geom_vline(xintercept = 453, color = "red", size = 0.5) +
  geom_vline(xintercept = 1789, color = "red", size = 0.5)
```

>>>>>>> d06e3acca5b9a2f9558f4ffba34549eeb7b22de2
# Section 4: Discussion - comment on advantages and disadvantages of the approach taken and
briefly discuss other approaches you could have taken instead.

# Appendix - contains any necessary technical and other material that would be good to include
but may make the main part difficult to read, particularly for a non-specialist/scientist. Sections
1-4 should be as accessible as possible to a non-statistician scientist/researcher in the relevant
area & should be written in a clear/concise manner.