---
title: "Assessing the Influence of DDE and PCBs on the Likelihood and Severity of Preterm Births"
author: "Phuc Nguyen, Joseph Lawson, Emily Gentles"
date: "1/23/2020"
output: bookdown::pdf_document2
toc: false
geometry: margin=0.7in
fontsize: 10pt
fig_caption: yes
references:
- id: brant1990
  title:  Assessing Proportionality in the Proportional Odds Model for Ordinal Logistic Regression
  author:
  - family: Brant
    given: Rollin
  container-title: Biometrics
  volume: 46
  URL: 'https://www.jstor.org/stable/pdf/2532457.pdf'
  publisher: International Biometric Society
  page: 1171-1178
  type: article-journal
  issued:
    year: 1990
    month: 12
- id: who
  title:  Preterm birth
  URL: 'https://www.who.int/news-room/fact-sheets/detail/preterm-birth'
  publisher: World Health Organization
  page: 1171-1178
  type: webpage
  issued:
    year: 2018
    month: 2   
- id: mccull1980
  title: Models for Ordinal Data
  author:
  - family: McCullagh
    given: Peter
  container-title: Journal of the Royal Statistical Society
  volume: 42
  URL: 'https://www.jstor.org/stable/2984952'
  publisher: Wiley
  page: 109-142
  type: article-journal
  issued:
    year: 1980
    month: 2
    
- id: cohn2011
  title:  Polychlorinated biphenyl (PCB) exposure in mothers and time to pregnancy in daughters
  author:
  - family: Cohn
    given: Barbara
  - family: Cirillo
    given: Piera
  - family: Sholtz
    given: Robert
  - family: Ferrara
    given: Assiamira
  - family: Park
    given: June-Soo
  - family: Schwingl
    given: Pamela
  container-title: Reproductive Toxicology
  volume: 31
  DOI: 'https://doi.org/10.1016/j.reprotox.2011.01.004'
  publisher: Elsevier
  page: 290-296
  type: article-journal
  issued:
    year: 2011
    month: 4
    
- id: cohn2003
  title: DDT and DDE exposure in mothers and time to pregnancy in daughter
  author:
  - family: Cohn
    given: Barbara
  - family: Cirillo
    given: Piera
  - family: wolff
    given: Mary
  - family: Schwingl
    given: Pamela
  - family: Cohen
    given: Richard
  - family: Sholtz
    given: Robert
  - family: Ferrara
    given: Assiamira
  - family: Christianson
    given: Roberta
  - family: van den Berg
    given: Barbara
  - family: Siiteri
    given: Pentti
  container-title: The Lancet
  volume: 361
  DOI: 'https://doi.org/10.1016/S0140-6736(03)13776-2'
  publisher: Elsevier
  page: 2205-2206
  type: article-journal
  issued:
    year: 2003
    month: 6
abstract: We examine a data set from the Collaborative Perinatal Project to assess how the risk of pre-term births are influenced by the mother's exposure to DDE and PCBs. We find positive evidence of an effect, using an ordinal logistic regression model to assess risk of different severities of pre-term births accounting for exposure to DDE, PCBs as well as potential socioeconomic and health-status confounders. Specifically, we find that a 10% increase in exposure to DDE is associated with a 1.8% increase in odds of having more premature birth, and a 10% increase in exposure to PCBs increases this odds by 0.6%. We discuss the advantages and drawbacks our methods as well as potential alternatives.

header-includes:
    \usepackage{float}
---

```{r setup, include=FALSE, warning = FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,fig.pos = 'H')
```

```{r, library, message=FALSE, results=FALSE, warning = FALSE}
library(tidyverse)
library(mice)
library(MASS)
library(GGally)
library(tidyverse)
library(mice)
library(nnet)
library(corrplot)
library(quantreg)
library(MASS)
library(knitr)
library(tigerstats)
library(ggplot2)
set.seed(43)
```

<!-- JOSEPH COMPUTATIONAL CODE BELOW -->

```{r, echo=FALSE, eval = TRUE, warning = FALSE, message = FALSE, cache=TRUE}
Long <- readRDS("Longnecker.rds")
longnecker_b <- Long
Long <- Long %>% mutate(albuminTested = ifelse(is.na(albumin), 0, 1)) %>% filter(gestational_age <= 44) %>% 
  filter(!is.na(pcb_028)) %>% dplyr::select(-albumin)



Long = Long %>% mutate(termCat = cut(gestational_age, c(breaks = c(0, 32, 37 ,45))), center = as.character(center),
                       pcb = pcb_028 + pcb_052 + pcb_074 + pcb_105 + pcb_118 + pcb_153 + pcb_170 +
                            pcb_138 + pcb_180 + pcb_194 + pcb_203,
                       # ed_norm = qnorm(score_education / 100),
                       # inc_norm = qnorm(score_education / 100),
                       # occ_norm = qnorm(score_occupation / 100),
                       dde.cut = cut(dde, breaks = 5),
                       pcb.cut = cut(pcb, breaks = 5),
                       gestord = factor(termCat, ordered = TRUE)
                       ) %>% filter(!is.na(pcb))


# ADD PRINCIPLE COMPONENTs
pcb_mat = as.matrix(Long %>% dplyr::select(contains("pcb_")))
min_pcb <- min(pcb_mat[pcb_mat > 0])
pcb_mat[pcb_mat == 0] <- min_pcb / 2
pcb_mat <- apply(pcb_mat, 2, log)
pcb.pr = prcomp(pcb_mat, center = T, scale = T)

Long = Long %>% mutate(logPCB1 = -pcb.pr$x[,1], logPCB2 = pcb.pr$x[,2], logPCB3 = pcb.pr$x[,3], logPCB4 = pcb.pr$x[,4],
                       logPCB5 = pcb.pr$x[,5], logPCB6 = pcb.pr$x[,6], logPCB7=pcb.pr$x[,7], logPCB8=pcb.pr$x[,8],
                       logPCB9 = pcb.pr$x[,9], logPCB10 = pcb.pr$x[,10], logPCB11 = pcb.pr$x[,11])

##

# Long_Score_NA_rm = Long %>% filter(!is.na(score_education + score_occupation + score_income))
# modordscore = polr(gestord ~ center + log(dde) + log(pcb) + score_education + score_occupation + score_education + log(triglycerides) + maternal_age + smoking_status + cholesterol + albuminTested, data = Long_Score_NA_rm)
# modordscorecomp = polr(gestord~ center+log(dde) + log(pcb)+log(triglycerides) + maternal_age + smoking_status + cholesterol + albuminTested, data = Long_Score_NA_rm)
# anova(modordscore, modordscorecomp)
# 
# modordcenter = polr(gestord~ center*log(dde) + center*log(pcb) + log(triglycerides) + maternal_age + smoking_status + cholesterol + albuminTested, data = Long)
# modord = polr(gestord~ center+log(dde) + log(pcb)+log(triglycerides)  + maternal_age + smoking_status + cholesterol + albuminTested, data = Long)
# summary(modord)
# anova(modord, modordcenter)


### BASE MODEL

Long.Base.Mod = polr(gestord~ center + log(dde) + logPCB1 + log(triglycerides)  + maternal_age + smoking_status + log(cholesterol) + 
                       albuminTested + race, data = Long,
                     Hess = T)
# summary(Long.Base.Mod)
Base.Confint = confint(Long.Base.Mod)
# Base.Confint
# knitr::kable(Base.Confint, "latex")

### PCA Analysis ###

Long.PCA.1.2.3 = polr(gestord~ center + log(dde) + logPCB1 + logPCB2 + logPCB3 + logPCB4 +
                        logPCB5 + logPCB6 + logPCB7 + logPCB8 + logPCB9 + logPCB10 +
                        logPCB11 + log(triglycerides)  + maternal_age + smoking_status + 
                        log(cholesterol) + albuminTested + race, data = Long, Hess = T)


# Long.PCA.logged = polr(gestord~ center + log(dde)*log(logPCB1) + log(triglycerides)  + maternal_age + smoking_status + cholesterol + albuminTested, data = Long)


PCA.anova = anova(Long.Base.Mod, Long.PCA.1.2.3)

## Second and third components add very little


#############################################
### MICE with ORDLOG - check score usefulness
#############################################

# impLong = mice(Long, printFlag = F)
impLong = mice(Long %>% mutate(ldde = log(dde), ltrig = log(triglycerides), lchol = log(cholesterol)) %>% 
                 dplyr::select(gestord,center, score_occupation, score_education, score_income, ldde, logPCB1,
                                 ltrig, maternal_age, smoking_status, lchol,
                                 albuminTested,race), printFlag = F)
imp.OL.Mod = with(impLong, polr(gestord ~ center + score_occupation +score_education +
                                  score_income + ldde + logPCB1 + 
                                    ltrig  + maternal_age + smoking_status + 
                                    lchol + albuminTested + race, Hess = T))

imp.OL.Mod.wo.score = with(impLong, polr(gestord ~ center  + ldde + logPCB1 + 
                                    ltrig  + maternal_age + smoking_status + 
                                    lchol + albuminTested + race, Hess = T))

score.p.value = pool.compare(imp.OL.Mod, imp.OL.Mod.wo.score)$pvalue

MICE.summary = summary(pool(imp.OL.Mod))

## Strong evidence that score does not add anything


##########################################
###### Check Center Interactions/inclusion
##########################################

center.inter.mod = polr(gestord~ center*log(dde) + center*logPCB1 + log(triglycerides)  + maternal_age + smoking_status + log(cholesterol) + albuminTested + race, data = Long, Hess = T)

Center.Inter.Anova = anova(center.inter.mod, Long.Base.Mod)

## Note that center interaction not significant

no.center.mod = polr(gestord~ log(dde) + logPCB1 + log(triglycerides)  + 
                       maternal_age + smoking_status + log(cholesterol) + albuminTested + race, data = Long, Hess = T)

Center.Inclusion.Anova = anova(no.center.mod, Long.Base.Mod)

# Strong evidence that including center is useful

```

```{r CHEM INTERACTIONS, echo=FALSE, eval = TRUE, warning = FALSE, message = FALSE, cache=TRUE}

##########################################
#### Check Chemical interactions
##########################################


## DDE PCA Interaction
Long.PCA.DDE.Inter.Mod = polr(gestord~ center + log(dde) * logPCB1 + log(triglycerides)  + maternal_age + 
                                smoking_status + log(cholesterol) + albuminTested + race, data = Long, Hess = T)

DDE.PCA.Inter.Anova = anova(Long.Base.Mod, Long.PCA.DDE.Inter.Mod)

# Not evidence that dde-pcb interact

## DDE/PCA Triglyceride Interaction
# Justified because these chemicals are fat-soluble.
Long.Trig.Inter.Mod = polr(gestord~ center + log(dde)*log(triglycerides) + logPCB1*log(triglycerides)  + 
                             maternal_age + smoking_status + log(cholesterol) + albuminTested + race, data = Long, Hess = T)

Trig.Inter.Anova = anova(Long.Base.Mod, Long.Trig.Inter.Mod)

```

```{r MATERNAL AGE POLY, echo=FALSE, eval = TRUE, warning = FALSE, message = FALSE, cache=TRUE}

##########################################
#### check maternal age poly #############
##########################################

Long.Age.Poly.Mod = polr(gestord~ center + log(dde) + logPCB1 + log(triglycerides)  + poly(maternal_age,2) + smoking_status + log(cholesterol) + albuminTested + race, data = Long, Hess = T)


Age.Poly.Anova = anova(Long.Base.Mod, Long.Age.Poly.Mod)


Long$MatAgeCat = cut(Long$maternal_age, 4)

MatAgeGestTable = rowPerc(xtabs(formula = ~ MatAgeCat + gestord, data = Long))
# plot(MatAgeGestTable[,1], type = "l")
# barplot(height = MatAgeGestTable[,1] / 100, xlab = "Age Bucket", ylab = "Severely Preterm Percentage", 
#         main = "Severely Preterm Probability vs Maternal Age")
## Conclude that the polynomial term is important.

```

```{r}


Long$white.or.not = as.factor(ifelse(Long$race == "white", "white", "non.white"))

Long.no.other.race.mod = polr(gestord ~ center + log(triglycerides) + log(dde) + logPCB1 +  maternal_age + smoking_status + log(cholesterol) + albuminTested + white.or.not, data = Long, Hess = T)

white.anova = anova(Long.no.other.race.mod, Long.Base.Mod)

```

```{r FINAL MODEL, echo=FALSE, eval = TRUE, warning = FALSE, message = FALSE, cache=TRUE}

##########################################
### FINAL MODEL ##########################
##########################################


Final.Mod = polr(gestord~ center + log(triglycerides) + log(dde) + logPCB1 +  poly(maternal_age,2) + smoking_status + log(cholesterol) + albuminTested + race, data = Long, Hess = T)

final.coef = coef(Final.Mod)
final.confint = confint(Final.Mod)

chem.effect.summary = cbind(final.coef[c("log(dde)","logPCB1")], final.confint[c("log(dde)", "logPCB1"),])
colnames(chem.effect.summary)[1] = "Coef Est"

Long$logdde = log(Long$dde)

Control.Mod = polr(gestord~ center + log(triglycerides) + poly(maternal_age,2) + smoking_status + log(cholesterol) + albuminTested + race, data = Long, Hess = T)

control.anova = anova(Control.Mod, Final.Mod)
control.p = control.anova$`Pr(Chi)`

### Brant test

# Long = Long %>% mutate(log_dde = log(dde), log_trig = log(triglycerides), log_chol = log(cholesterol), maternal_age_2 = maternal_age^2)
# 
# Brant.Mod = polr(gestord~ center + log_trig + log_dde + logPCB1 + maternal_age + maternal_age_2 + smoking_status + log_chol + albuminTested + race, data = Long, Hess = T)
# 
# require(brant)
# mod.brant = brant(Brant.Mod)


```

<!-- END JOSEPH CODE -->

<!-- ## Abstract/Summary - summarize the problem and the main findings of your project. -->

<!-- ## Section 1: Introduction - provide an initial description of the data, study in which it was collected and focus of the study/analysis/questions of interest. -->

# Introduction

  The data used in this analysis is a subsample of the data collected in the National Collaborative Perinatal Project (CPP), a program which enrolled women during pregnancy and followed their pregnancy outcomes. The pregnancy outcome of interest in this analysis is premature delivery, which is typically defined as having gestational age at delivery of 37 weeks or less. The project collected demographic information including the mother's race, age, education, occupation as well as a lifestyle factor that is smoking status. The subsample contains 2380 women for whom additional assaying of serum samples was perfromed and the concentration doses for DDE and PCBs as well as cholesterol and triglycerides was recorded. DDE and PCBs are of particular interest as they are breakdown products of pesticides that bioaccumulate in fatty deposits of organisms and are thought to have an impact on human reproduction [see @cohn2003; also @cohn2011]. The goal of this analysis is to assess how exposure to DDE and PCBs relates to the risk of premature delivery. 

<!-- # Section 2: Materials & Methods - describe the statistical approach/approaches used in the analysis, providing references as appropriate; do not present results in this section but just the methods used. You can carry out your analysis with multiple methods. -->

# Materials & Methods

In order to assess how exposure to DDE and PCBs relates to the risk of premature delivery, we investigate the associations between the concentration of DDE and PCBs and categories of preterm birth, accounting for other potential confounders. Specifically, we define categories of preterm birth as severely premature for having gestational ages of less than 33 week, premature for having gestational ages of less than 38 weeks, and full- to late-term for having gestational ages of between 38 and greater based on the guidelines provided by the World Health Organization [@who]. The natural order of preterm birth categories lead us to employ ordinal logistic regression as our modeling method [@mccull1980]. This method provides interpretations in terms of change in risk of more premature birth as exposure to DDE and PCBs increases. This structure also addresses the non-normality, heavy-tailedness of the response which impedes the use of standard linear regression. The model has the following structural form:

$$ logit(Y_i\leq j) = \alpha_j - \beta^T X_i $$
From which one derives:

$$ P(Y_i\leq j) = \frac{e^{\alpha_j - \beta^T X_i}}{1+e^{\alpha_j-\beta^T X_i}} $$



In the above, $Y_i$ is the response category, $j$ is some possible level of $Y_i$, $\alpha_j$, is a constant associated with level $j$, $\beta$ is a vector of coefficients shared across all levels, and $X_i$ is the vector of predictors for the $i^{th}$ observation. Note that unlike multinomial regression, in which one obtains $P(Y_i = j)$, we have an inequality instead. The phrase "proportional odds" comes from the assumption under the model that the odds ratio of $Y_i\leq j$ vs $Y_k\leq j$ is constant across the range of $j$.


The reasonable use ordinal logstic regression relies on several assumptions: ordered categorical response variable, proportional odds, lack of multi-collinearity, and predictors are either categorical, ordinal, or continuous. We perform a Principal Component Analysis (PCA) of the PCB predictors, which exhibit relatively high correlations (\ref{fig:gpairs} (right)), to produce new uncorrelated variables, guarenteeing the multi-collinearity assumption. The proportional odds assumption may be tested via the Brant test for which we obtain satisfactory results [@brant1990]. We note that "Score" variables have a high degree of missingness and use the MICE package in R to impute these missing data points. We remove one observation with missing PCB data. We also include other potential confounders such as blood cholesterol/triglyceride levels, the center where these measurements were taken and metrics capturing socioeconomic status and lifestyle in the model. Finally, we perform model selection using F-tests.

<!-- FROM SLIDE SHOW BELOW -->

<!-- - Ordinal Logistic Regression with Term, Preterm, and Severely Preterm Gest. Categories -->
<!--   + Useful interpretation in terms of risk -->
<!--   + Uses naturally ordinal structure -->
<!--   + Differentiates between different severities -->
<!--   + Addresses non-normality of response -->
<!-- - Keep obs with Gest. Age 44 or less -->
<!-- - Impute score data with MICE to check usefulness -->
<!-- - Remove obs with missing PCE value -->
<!-- - Include blood cholesterol/triglyceride levels, as well as center and SES/Lifestyle metrics -->

<!-- ## Section 3: Results - describe the results of the analysis approaches described in Section 2. It -->
<!-- may be useful to break this into subsections, such as: -->
<!-- – 3.1 Exploratory Data Analysis -->
<!-- – 3.2 Main Results -->
<!-- – 3.3 Sensitivity Analysis -->
<!-- Make sure to very clearly address the questions of interest! -->
# Results

## Exploratory Data Analysis

Many of the variables in the dataset exhibit a variety of issues such as measurement errors, zero-inflation, multi-collinearity and missingness. Gestational age potentially contain measurement errors since its values range from 27 to 90 (Figure \ref{fig:gesthist} left). We address this issue by keeping observations with 44 weeks or less in gestational age, a range plausible in reality.

We also see an inflation at zeros in the PCB variables due to the limitations of measurement equipment (Figure \ref{fig:gesthist} right). Consequently, we add half of the smallest value of PCB measurements to the zero and log-transform these variables. As seen in Figure \ref{fig:gpairs} (right), the PCB variables are highly correlated. Applying PCA to these variables addresses the issues. Among the predictors, only DDE and the first principal component of PCBs, triglycerides and cholesterol are mildly correlated (Figure \ref{fig:gpairs} left).

Missigness is most significantly present in the variable albumin for which less than 10% of women had their levels recorded. Figure \ref{fig:bivartrend} shows that being tested for albumin seems to associate with slightly larger gestational age. Bivariate plots show associations between premature birth and predictors such as race, centers, a quadratic function of maternal age, concentration of DDE and cholesterol (Figure \ref{fig:bivartrend}).

```{r, clean-data}
longnecker_base <- readRDS(file = "Longnecker.rds")

# Keep gestational age less than 45
# Remove 1 row of pcb_* with NA
# Dichotomize gestational age into 4 categories
# Create indicator of albumin
# Change center to categorical
# Combine race black and race other
gestational_age_cutoff <- 45
longnecker <- longnecker_base %>%
  filter(gestational_age < gestational_age_cutoff) %>%
  filter(!is.na(pcb_028)) %>%
  mutate(has_albumin = 1 * !is.na(albumin)) %>%
  mutate(center = as.character(center)) %>%
  mutate(gestational_cat = as.factor(ifelse(gestational_age <= 32, 1, 
                                            ifelse(gestational_age <= 37 & gestational_age > 32, 2, 3))))

# Log transform pcb_* after adding a small number to 0 values
pcb_mat <- longnecker %>%
                    dplyr::select(contains("pcb")) %>%
                    data.matrix()
min_pcb <- min(pcb_mat[pcb_mat > 0])
pcb_mat[pcb_mat == 0] <- min_pcb / 2
pcb_mat <- apply(pcb_mat, 2, log)
# Summarize pcb_* with PCA
pcb_pca <- prcomp(pcb_mat, scale = TRUE, center = TRUE, retx = TRUE)
# To see loading of each variable, uncomment the below:
#pcb_pca$rotation[, 1] * (pcb_mat[1, ] - pcb_pca$center)/ pcb_pca$scale
# To see variation explained by each PC, uncomment the below:
# pcb_pca$sdev^2 # == > two PCs are enough
longnecker["pcb_pca1"] <- pcb_pca$x[, 1]
longnecker["pcb_pca2"] <-  pcb_pca$x[, 2]

# Log transform these variables
longnecker["log_dde"] <- log(longnecker$dde)
longnecker["log_tri"] <- log(longnecker$triglycerides)
longnecker["log_chol"] <- log(longnecker$cholesterol)
```

```{r gpairs, message=FALSE, fig.show="hold", out.width="50%", fig.cap="\\label{fig:gpairs}Correlations between predictors. First principal component of log(PCB) and log(DDE) are mildly negatively correlated (-0.3), and triglycerides and cholesterol are mildly positively correlated (0.3) (left). The PCB variables are all highly correlated (right)."}
ggpairs(longnecker %>% dplyr::select(gestational_cat, log_dde, pcb_pca1, log_tri, cholesterol, maternal_age))
longnecker %>%
  dplyr::select(contains("pcb")) %>%
  data.matrix() %>%
  cor() %>%
  round(1) %>%
  reshape2::melt() %>%
  ggplot(aes(x=Var1, y=Var2, fill=value)) + 
  geom_tile() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + xlab("") + ylab("")
```

```{r gesthist, out.width="50%", fig.show="hold", fig.cap="\\label{fig:gesthist}Gestational age ranges from 27 to 90 weeks (left). Distributions of some PCBs measurements"}
ggplot(longnecker_base, aes(x = gestational_age)) + geom_histogram(bins = 50) + geom_vline(aes(xintercept = 45),color="blue", linetype="dashed", size=1)
longnecker %>%
  dplyr::select(pcb_170, pcb_138, pcb_203) %>%
  gather(key = "var", value = "val") %>%
  ggplot(aes(x = val)) + geom_histogram(bins=30) + facet_wrap(~var)
```

```{r bivartrend, out.width="33%", fig.show="hold", fig.cap="\\label{fig:bivartrend}Bivariate plots of gestational categeries (1: serverely-premature, 2: premature, 3: normal-late) against predictors"}
ggplot(longnecker, aes(fill = race, x = gestational_cat)) + geom_bar()
ggplot(longnecker, aes(y = pcb_pca1, x = gestational_cat)) + geom_boxplot()
ggplot(longnecker, aes(y = log_dde, x = gestational_cat)) + geom_boxplot()
ggplot(longnecker, aes(y = triglycerides, x = gestational_cat)) + geom_boxplot()
ggplot(longnecker, aes(y = cholesterol, x = gestational_cat)) + geom_boxplot()
longnecker %>%
  mutate(smoker = as.character(smoking_status)) %>%
  ggplot(aes(fill = smoker, x = gestational_cat)) + geom_bar()
ggplot(longnecker, aes(fill = gestational_cat, x = center)) + geom_bar()
longnecker %>%
  dplyr::select(contains("score"), gestational_cat) %>%
  gather(key = "variable", value = "value", - gestational_cat) %>%
  ggplot(aes(colour = gestational_cat, y = value, x = variable)) + geom_boxplot()
Long$MatAgeCat = cut(Long$maternal_age, 4)
MatAgeGestTable = rowPerc(xtabs(formula = ~ MatAgeCat + gestord, data = Long))
barplot(height = MatAgeGestTable[,1] / 100, xlab = "Age Bucket", ylab = "Severely Preterm Percentage",
        main = "Severely Preterm Probability vs Maternal Age")
longnecker %>%
  mutate(has_albumin = as.factor(has_albumin)) %>%
  ggplot(aes(y = gestational_age, x = has_albumin)) + geom_boxplot() + xlab("Albumin tested indicator")
```

<!-- ### Ordinal Logistic Regression -->

```{r, ord}
ord_mod <- polr(gestational_cat ~ race +
                 log_dde + 
                 pcb_pca1 +
                 has_albumin +
                 poly(maternal_age, 2)+ 
                 log_tri +
                 log_chol +
                 smoking_status +
                 center, 
               data = longnecker,
               Hess=TRUE)
p_ord <- pnorm(abs(coef(summary(ord_mod))[, "t value"]), lower.tail = FALSE) * 2
ord_mod_summary = cbind(summary(ord_mod)$coefficients, "p value" = p_ord)
```

## Main Results

We perform F-tests to determine the appropriate structure and complexity of our model. 

* Indication (p=`r round(PCA.anova[2,"Pr(Chi)"],2) `) that the first principle component of the pcb_* values is sufficient.
* Indication (p=`r round(score.p.value, 2)`) against including Score Variables (post imputation)
* Indication (p=`r round(Center.Inter.Anova[2, "Pr(Chi)"],2) `) against including Center interactions with DDE/PCB
* Indication (p=`r round(Center.Inclusion.Anova[2, "Pr(Chi)"],2) `) for including Center as variable
* Indication (p=`r round(DDE.PCA.Inter.Anova[2, "Pr(Chi)"],2) `) against PCB-DDE interaction effect
* Indication (p=`r round(Trig.Inter.Anova[2, "Pr(Chi)"],2) `) (weakly) against Triglyceride interaction with PCE/DDE
* Indication (p=`r round(Age.Poly.Anova[2, "Pr(Chi)"],2) `) for inclusion of quadratic term in maternal age
* Strong indication that the indicator of testing for Albumin is associated with longer gestational period on the margin.

Taking into account these results, we express our final model form below in R-standard notation:

\begin{align} GestCategory &\sim Center+log(dde)+logPCB1 \\ &+log(trigl.)+Poly(MaternalAge,2) \\&+Smoking+log(Cholest.)+AlbuminTested \\  &+race\end{align} 

Testing inclusion of both DDE and PCB vs. control model gives p=`r round(control.p[2],3)`. This indicates that DDE and PCB improve the explanatory power of pre-term birth risk model. Predictive check for the model is seen in Figure \ref{fig: predcheck} and marginal fit showed in Figure \ref{fig:predmar}. Marginally in the full model each is at the edge of significance (at the 5% level):
<!-- - Note the overlap of the confidence intervals with 0. -->

```{r}
library(kableExtra)
knitr::kable(as.data.frame(round(1*chem.effect.summary, 3)), caption = "Coefficient Estimates for Target Chemicals") %>% kable_styling(latex_options = "HOLD_position")
```


```{r}
inter_table <- chem.effect.summary
rownames(inter_table) <- c("DDE", "PCB1")
resp = as.data.frame(round(exp(.1*inter_table),3))
knitr::kable(resp, caption = "Multiplier of odds ratio for 10 percent increase in exposure") %>% kable_styling(latex_options = "HOLD_position")
```

These estimates suggests evidence for an association between exposure to DDE and PCBs and an increase in risk of premature birth. We can interprete these coefficents as follows:

- For 10% increase in DDE exposure, the odds of having more normal gestational age decreases by `r round(100 - 100 * exp(.1*inter_table[1,1]),1)`%.
- For 10% increase in the first principal components of PCBs, the odds of having more normal gestational age decreases by `r round(100 - 100 * exp(.1*inter_table[2,1]),1)`%. See Figure \@ref(fig:Results) below for a pictoral representation of fitted dose response probabilities across gestational term severities. 

<!-- https://stats.idre.ucla.edu/r/dae/ordinal-logistic-regression/ -->
<!-- https://stats.idre.ucla.edu/other/mult-pkg/faq/general/faqhow-do-i-interpret-a-regression-model-when-some-variables-are-log-transformed/ -->

```{r Results, fig.width=8, fig.cap="\\label{fig:Results}Predicted Dose Responses as Probabilities for Log DDE and Log PCB1"}
probabilities <- predict(Final.Mod, type = "probs")
predictors <- c("logdde", "logPCB1","severe_premature", "premature", "normal"
                )
Long %>%
  mutate(severe_premature = probabilities[,1],
         premature = probabilities[, 2],
         normal = probabilities[, 3]
         ) %>%
  dplyr::select(predictors) %>%
  gather(key = "predictors", value = "predictor.value", 
         -severe_premature, -premature, -normal) %>%
  gather(key = "categories", value = "probability", -predictors, -predictor.value) %>%
  ggplot(aes(probability, predictor.value)) +
  geom_point(size = 0.5, alpha = 0.5) +
  geom_smooth(method = "loess") + 
  theme_bw() + 
  facet_grid(cols = vars(categories), rows = vars(predictors), scales = "free") +
  ylab("DDE or PCB1 Value (log scale)")
```

## Sensitivity Analysis

We refit the above model for different cutoffs defining severely-premature, premature, and normal-late birth categories and plot the 95% CI's of the significant predictors in Figure \ref{fig:sensitivity-cutoff}. The plot shows that the CIs are relatively robust to reasonable change in cutoffs of the response categories.

```{r sensitivity-cutoff, cache=TRUE, fig.cap="\\label{fig:sensitivity-cutoff}95% CI of significant predictors for different cutoffs defining preterm birth categories"}
ci <- confint(ord_mod)
ci_df <- data.frame(ci) %>%
  rownames_to_column() %>%
  mutate(model = "32-37 (main)")
cutoffs <- list(c(31, 36), c(33, 38), c(31, 38))
for (cutoff in cutoffs) {
  longnecker_test <- longnecker %>%
    mutate(gestational_cat = as.factor(ifelse(gestational_age <= cutoff[1], 1, 
                                       ifelse(gestational_age <= cutoff[2] & gestational_age > cutoff[1], 2, 3))))
  test_mod <- polr(gestational_cat ~ race +
                 log_dde + 
                 pcb_pca1 +
                 has_albumin +
                 poly(maternal_age, 2) + 
                 log_tri +
                 log_chol +
                 smoking_status +
                 center, 
               data = longnecker_test,
               Hess=TRUE)
  ci <- confint(test_mod)
  df <- data.frame(ci) %>% rownames_to_column() %>% mutate(model = paste0(cutoff[1], "-", cutoff[2]))
  ci_df <- rbind(ci_df, df)
}
pd <- position_dodge(0.2)
ci_df["mean"] <- (ci_df$X2.5.. + ci_df$X97.5..)/2
ggplot(ci_df %>% filter(rowname %in% c("raceblack", "log_dde", "pcb_pca1", "has_albumin", "log_tri", "log_chol")), 
       aes(y=mean, x=rowname, colour=model)) + 
  geom_errorbar(aes(ymin=X2.5.., ymax=X97.5..), width=.2, position=pd) +
  labs(colour = "Preterm Birth Cutoffs")
```

# Discussion 

<!-- - comment on advantages and disadvantages of the approach taken and briefly discuss other approaches you could have taken instead. -->

A key advantage of this approach is that its interpretation is naturally suited to the problem at hand. Our goal is to assess risks of pre-term delivery resulting from increased exposure to DDE and PCBs. An ordinal logistic regression model provides probability of different categories of preterm birth and their natural order. One disadvantage of this model is that it cannot capture different effects of the exposure on different severity of preterm birth, if such relationship exists. Another disadvantage is the necessity of cut points to define preterm birth categories. In one sense, the existence of certain natural cut points in fetal development, as provided by various reputable organizations as WHO, may make this type of categorization seem sensible. On the other hand, categorizing continuous data will always possess a degree of arbitrarity, and it may risk masking different dynamics contained within categories or across categories. Alternative modeling frameworks that address these limitations include quantile regression, generalized additive model, B-spline or density regression.

We find that the first principle component is sufficient. As the weights in the first principle component are all of the same sign and of broadly similar magnitude, we could have perhaps simply taken the sum of the PCBs predictors as an alternative methodology. This would improve the interpretability of the summary predictor. 

<!-- # Appendix - contains any necessary technical and other material that would be good to include but may make the main part difficult to read, particularly for a non-specialist/scientist. Sections 1-4 should be as accessible as possible to a non-statistician scientist/researcher in the relevant area & should be written in a clear/concise manner. -->

# Appendix

## Final Model Summary Output


```{r}
 summary(Final.Mod) 
```


## ANOVA Tests

```{r}
library(kableExtra)
knitr::kable(PCA.anova %>% dplyr::select(-Model), caption = "PCA Anova") %>% kable_styling(latex_options = "HOLD_position")

knitr::kable(Center.Inter.Anova %>% dplyr::select(-Model), caption = "Center Interaction Anova")%>% kable_styling(latex_options = "HOLD_position")

knitr::kable(Center.Inclusion.Anova %>% dplyr::select(-Model), caption = "Center Inclusion Anova")%>% kable_styling(latex_options = "HOLD_position")

knitr::kable(DDE.PCA.Inter.Anova %>% dplyr::select(-Model), caption = "DDE-PCB1 Interaction Anova")%>% kable_styling(latex_options = "HOLD_position")

knitr::kable(Trig.Inter.Anova %>% dplyr::select(-Model), caption="Triglyceride-DDE/PCB Interaction Anova")%>% kable_styling(latex_options = "HOLD_position")

knitr::kable(Age.Poly.Anova %>% dplyr::select(-Model), caption = "Quadratic Maternal Age Anova")%>% kable_styling(latex_options = "HOLD_position")

knitr::kable(control.anova %>% dplyr::select(-Model), caption = "DDE/PCB vs Control Anova")%>% kable_styling(latex_options = "HOLD_position")
```

## Proportional Odds Assumption

Below is the output for the Brant test mentioned above:
<!-- JUST A BLOG POST: (from this post here: https://medium.com/evangelinelee/brant-test-for-proportional-odds-in-r-b0b373a93aa2).  -->

```{r brant, results='hide'}
library(brant)
test_mod <- polr(gestational_cat ~ race +
                 log_dde + 
                 pcb_pca1 +
                 has_albumin +
                 maternal_age + 
                 log_tri +
                 log_chol +
                 smoking_status +
                 center, 
               data = longnecker,
               Hess=TRUE)
br.test = brant(test_mod)
```

```{r}
knitr::kable(br.test, caption = "Brant test for proportional odds") %>% kable_styling(latex_options = "HOLD_position")
```

Note that we have a significant p-value for Race:Other; otherwise there are robust results indicating that the proportional odds assumption is not unreasonable. We note that because the Race:Other category is quite small as a percentage of the whole data set, it is likely to be acceptable to proceed with this modeling procedure. On a more general note, the proportional odds assumption is evident in the following computation:

$$ \frac{odds(y_i\leq j)}{odds(y_k\leq j)} = \frac{e^{\alpha_j - \beta^T x_i}}{e^{\alpha_j - \beta^T x_k}} = e^{\beta^T(x_k-x_i)} $$
Note how in the above this ratio is a constant over the range of $j$. As is fairly evident, this is a rather strict requirement to place on one's data.

## Ordinal Logistic Model

We perform some short calculations to demonstrate how one solve for probabilities of individual ordinal values from the modeled output of the model. We have:

\begin{align*} logit(y_i\leq j) &= \alpha_j - \beta^T x_i \Rightarrow \\
  P(y_i\leq j) &= \frac{e^{\alpha_j - \beta^T x_i}}{1+e^{\alpha_j - \beta^T x_i}} \\
  P(y_i = j) &= P(y_i \leq j) - P(y_i\leq j-1) = \frac{e^{\alpha_j - \beta^T x_i}}{1+e^{\alpha_j - \beta^T x_i}} - \frac{e^{\alpha_{j-1} - \beta^T x_i}}{1+e^{\alpha_{j-1} - \beta^T x_i}} \\
  &= \frac{e^{\alpha_j-\beta^T x_i} - e^{\alpha_{j-1}-\beta^T x_i}}{(1+e^{\alpha_j-\beta^T x_i})(1+e^{\alpha_{j-1} - \beta^T x_i})}
\end{align*}

## Model Checking

See Figure \ref{fig:predcheck} and Figure \ref{fig:predmar}

```{r predcheck, cache=TRUE , fig.cap="\\label{fig:predcheck}Predictive check for distribution of count of each preterm birth category"}
  y <- longnecker$gestational_cat
  n <- 200
  y_hat <- matrix(NA, nrow = length(y), ncol = n)
  for (i in 1:n) {
    y_hat[, i] <- apply(probabilities, 1, function(x) {which(rmultinom(1, 1, x) == 1)})
  }
  
  # Distribution of count of each category in the simulated data
  count_severe <- apply(y_hat, 2, function(x) {sum(x == 1)})
  count_mild <- apply(y_hat, 2, function(x) {sum(x == 2)})
  count_normal <- apply(y_hat, 2, function(x) {sum(x == 3)})
  sens_data <- data.frame(severe = count_severe, mild = count_mild, normal = count_normal) %>%
    gather(key = "categories", value = "count")
  ggplot(sens_data, aes(x = count, fill = categories)) + 
    geom_histogram(bins = 200) +
    geom_vline(xintercept = table(y)[1], color = "red", size = 0.5) +
    geom_vline(xintercept = table(y)[2], color = "red", size = 0.5) +
    geom_vline(xintercept = table(y)[3], color = "red", size = 0.5)
```

```{r predmar,  fig.cap="\\label{fig:predmar}Visualizing predicted logit against covariates"}
# Tutorial from http://www.sthda.com/english/articles/36-classification-methods-essentials/148-logistic-regression-assumptions-and-diagnostics-in-r/
probabilities <- predict(ord_mod, type = "probs")
predictors <- c("log_dde", 
                 "pcb_pca1",
                 "log_tri",
                 "cholesterol",
                 "severe_premature", "premature", "normal"
                )
longnecker %>%
  mutate(severe_premature = log(probabilities[, 1]/(1-probabilities[, 1])),
         premature = log(probabilities[, 2]/(1-probabilities[, 2])),
         normal = log(probabilities[, 3]/(1-probabilities[, 3]))
         ) %>%
  dplyr::select(predictors) %>%
  gather(key = "predictors", value = "predictor.value", 
         -severe_premature, -premature, -normal) %>%
  gather(key = "categories", value = "logit.value", -predictors, -predictor.value) %>%
  ggplot(aes(logit.value, predictor.value))+
  geom_point(size = 0.5, alpha = 0.5) +
  geom_smooth(method = "loess") + 
  theme_bw() + 
  facet_grid(cols = vars(categories), rows = vars(predictors), scales = "free")
```

# Citations
<!-- cut offs -->
<!-- https://www.who.int/news-room/fact-sheets/detail/preterm-birth -->

<!-- albumin -->
<!-- https://account.allinahealth.org/library/content/1/3480 -->
<!-- https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4665611/ -->